---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# surveycore

<!-- badges: start -->
[![R-CMD-check](https://github.com/JDenn0514/surveycore/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/JDenn0514/surveycore/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/JDenn0514/surveycore/graph/badge.svg)](https://app.codecov.io/gh/JDenn0514/surveycore)
<!-- badges: end -->

**surveycore** is the foundation of the surveyverse ecosystem — a modern,
tidyverse-compatible replacement for the `survey` and `srvyr` packages in R.

It provides S7-based survey design objects with:

- A tidy-select interface (`ids = c(psu, ssu)`, no formula syntax)
- Automatic preservation of haven-style variable labels and value labels
- Exact variance estimation (Taylor linearization, replicate weights,
  two-phase designs)
- Seamless conversion to and from `survey::svydesign` and `srvyr::tbl_svy`

## Installation

```r
# From CRAN:
install.packages("surveycore")

# Development version from GitHub:
# install.packages("pak")
pak::pak("JDenn0514/surveycore")
```

## What Phase 0 provides

- **S7 survey objects**: `survey_taylor`, `survey_replicate`, `survey_twophase`
- **Constructors**: `as_survey()`, `as_survey_rep()`, `as_survey_twophase()`
- **Metadata system**: `set_var_label()`, `set_val_labels()`, `extract_var_label()`,
  `extract_val_labels()` — with automatic haven attribute import
- **Variance estimation**: `get_means()`, `get_totals()` — Taylor linearization
  and replicate weights
- **Design utilities**: `update_design()`, `as_svydesign()`, `from_svydesign()`,
  `as_tbl_svy()`, `from_tbl_svy()`

## Basic usage

```{r example}
library(surveycore)

# ── Simple SRS design ──────────────────────────────────────────────────────────
set.seed(42)
df <- data.frame(
  psu     = rep(1:10, each = 10),
  strata  = rep(c("A", "B"), each = 50),
  weight  = runif(100, 0.5, 2),
  income  = rnorm(100, 50000, 10000),
  age     = sample(18:80, 100, replace = TRUE)
)

d <- as_survey(df, ids = psu, weights = weight, strata = strata, nest = TRUE)
d

# ── Weighted mean and total ────────────────────────────────────────────────────
get_means(d, income)
get_totals(d, income)
```

## Complex survey designs

```{r complex}
# ── Replicate weights (BRR) ───────────────────────────────────────────────────
df_rep <- data.frame(
  y    = rnorm(20),
  wt   = runif(20, 1, 3),
  rep1 = runif(20, 0.5, 2),
  rep2 = runif(20, 0.5, 2),
  rep3 = runif(20, 0.5, 2),
  rep4 = runif(20, 0.5, 2)
)

d_rep <- as_survey_rep(
  df_rep,
  weights    = wt,
  repweights = starts_with("rep"),
  type       = "BRR"
)
d_rep
```

## Variable labels

surveycore preserves haven-style labels automatically when reading `.xpt` or
`.sav` files. You can also set labels manually:

```{r labels}
d2 <- set_var_label(d, income, "Annual household income (USD)")
d2 <- set_var_label(d2, age, "Respondent age in years")

extract_var_label(d2, income)
extract_var_label(d2, age)
```

## Conversion to/from survey and srvyr

```{r conversion, eval = FALSE}
# To survey::svydesign
svy <- as_svydesign(d)
class(svy)
#> [1] "survey.design2" "survey.design"

# Back to surveycore
d_rt <- from_svydesign(svy)
d_rt
```

## What's coming

- **Phase 0.5** — `filter()`, `select()`, `mutate()`, `group_by()` verbs
  via the `surveytidy` package
- **Phase 1** — `get_freqs()`, `get_quantiles()`, `get_ratios()`,
  `get_diffs()`, `get_corr()`
- **Phase 2** — `survey_glm()`, `survey_lm()` regression
- **Phase 2.5** — Calibration / post-stratification via `surveyweights`

## License

GPL-3. Variance estimation code vendored from the
[`survey`](https://cran.r-project.org/package=survey) package (Thomas Lumley,
GPL-2/GPL-3) — see `VENDORED.md` for full attribution.
