---
title: "Getting Started with surveycore"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Getting Started with surveycore}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(surveycore)
knitr::opts_chunk$set(
  comment = "#>"
)
```

`surveycore` gives you a (mostly) complete workflow for survey data analysis.
This vignette is designed to give you a quick overview of the main functionality 
present in this package. It is comprised of three main sections:

1. Creating survey objects 

2. Conducting simple analysis

3. Managing the metadata

**Quick PSA before jumping in:**

`surveycore` was built as an alternative to `survey` and `srvyr`. However, the
code powering the variance estimation and analysis is vendored from the `survey` 
package. Everything aspect of this package that calculates anything has been 
tested to ensure it provides the same numerical results. Without Thomas Lumley's 
work on that package, surveycore would not be possible,

---

## Create the survey object

The first step when conducting survey analysis is creating a right survey object 
where we specify the sampling design, weights, and whatever other information is
need. Without this information, point estimates may be biased and standard errors 
are almost certainly wrong [@lumley2010; @lohr2022].

Fortunately, we don't have to calculate that uncertainty ourselves! That's what
the survey objects are for. They tell the analysis functions how to conduct 
its analysis so they can properly take into account the variance and bias from
the survey design. 

`surveycore` has five different survey object constructors: 

1. `as_survey()`

2. `as_survey_rep()`

3. `as_survey_srs()`

4. `as_survey_calibrated()`

5. `as_survey_twophase()`

Rather going into detail on each constructor, I'm just going to provide a quick
overview of each. For more information visit `vignette("creating-survey-objects")`

### `as_survey()` 

You want to use this if you used a probability sample and the data you have has 
cluster IDs, strata, and/or design weights. There are several


### `as_survey_rep()`
Use this when the data you have comes with pre-built replicate weight columns 
like `repwt_1`, `repwt_2`. 

3. `as_survey_srs()` - Use this when your data is a pure SRS meaning each respondent has equal probability of selection and there are no clustering nor strata.

4. `


```{r design}
# NHANES: stratified cluster design using Taylor series linearization.
# Subset to MEC exam participants before using the exam weight (wtmec2yr).
nhanes_exam <- nhanes_2017[nhanes_2017$ridstatr == 2, ]

svy <- as_survey(
  nhanes_exam,
  ids = sdmvpsu, # cluster ID
  strata = sdmvstra, # stratum
  weights = wtmec2yr, # examination weight
  nest = TRUE # PSU IDs are locally unique within strata
)
svy
```

`as_survey()` is the right constructor for probability surveys with cluster
and stratum information. `nest = TRUE` is needed here because NHANES assigns
PSU IDs (1 and 2) locally within each stratum, not globally across the
dataset. Use `vignette("creating-survey-objects")` to find the right
constructor for your survey.

---

## 2. Frequency tables — `get_freqs()`

`get_freqs()` estimates the weighted distribution of one or more categorical
variables. Pass a bare column name as the second argument.

```{r freqs-basic}
# Distribution of sex (1 = Male, 2 = Female)
get_freqs(svy, riagendr)
```

The result is a `survey_freqs` tibble. `pct` is the survey-weighted percentage;
`n` is the unweighted cell count.

### Confidence intervals and margins of error

Add uncertainty columns with the `variance` argument. Any combination of
`"se"`, `"ci"`, `"moe"`, `"var"`, `"cv"`, or `"deff"` is supported.

```{r freqs-variance}
get_freqs(svy, riagendr, variance = c("ci", "moe"))
```

`ci_low` / `ci_high` are 95% Wilson-score confidence interval bounds.
`moe` is the margin of error at the specified confidence level
($\pm \mathrm{MOE}$ gives the half-width of the CI).

### Multiple variables at once

Pass a `c()` of variable names to get a stacked long-format frequency table.

```{r freqs-multi}
# Race/ethnicity (ridreth3) and sex (riagendr) in one call
get_freqs(svy, c(riagendr, ridreth3))
```

The `name` column identifies which variable each row belongs to; `value`
holds the response code.

### Including `NA` as a level

By default `na.rm = TRUE` drops missing values before computing
percentages. Set `na.rm = FALSE` to include `NA` as a level:

```{r freqs-na}
get_freqs(svy, dmdeduc2, na.rm = FALSE)
```

---

## 3. Weighted means — `get_means()`

`get_means()` estimates the survey-weighted mean of a continuous variable.

```{r means-basic}
# Mean systolic blood pressure
get_means(svy, bpxsy1)
```

The default output includes the mean and 95% confidence interval. Supply
`variance = "se"` to replace the CI with a standard error, or combine
options:

```{r means-variance}
get_means(svy, bpxsy1, variance = c("se", "ci", "deff"))
```

`deff` is the design effect — the ratio of the complex-design variance to the
variance under simple random sampling. Values above 1 indicate that clustering
and/or stratification inflates sampling uncertainty compared to SRS.

---

## 4. Population totals — `get_totals()`

`get_totals()` estimates the weighted sum: the total count or aggregate for
the target population.

### Population size

Called without a variable argument, `get_totals()` estimates the total
number of persons (or units) the sample represents — the sum of the survey
weights.

```{r totals-pop}
# Estimated U.S. civilian non-institutionalized population, 2017-2018 (exam)
get_totals(svy)
```

### Variable total

Pass a numeric column to estimate its population total.

```{r totals-var}
# Estimated total pulse (bpxpls) across the population
get_totals(svy, bpxpls)
```

---

## 5. Subgroup analysis — the `group` argument

Every analysis function accepts a `group` argument for computing estimates
separately within levels of a categorical variable. Pass a bare column name
or a `c()` of names.

```{r group-means}
# Mean systolic blood pressure by sex
get_means(svy, bpxsy1, group = riagendr)
```

```{r group-totals}
# Population size by race/ethnicity
get_totals(svy, group = ridreth3)
```

```{r group-freqs}
# Education distribution by sex
get_freqs(svy, dmdeduc2, group = riagendr)
```

Rows where the grouping variable is `NA` are excluded from all groups and do
not appear in the output. Responses within each group sum to 100% for
`get_freqs()`.

---

## 6. Ratio estimation — `get_ratios()`

`get_ratios()` estimates the ratio of two weighted totals:

$$\hat{R} = \frac{\sum_i w_i \, y_i}{\sum_i w_i \, x_i}$$

Variance is estimated via the delta method (linearization), equivalent to
`survey::svyratio()` [@lumley2010].

```{r ratios}
# Ratio of systolic to diastolic blood pressure
get_ratios(svy, numerator = bpxsy1, denominator = bpxdi1)
```

Ratios are useful when you want an estimate that is invariant to the scale of
the weights — for example, wages per hour, spending per household member, or
disease prevalence ratios.

```{r ratios-grouped}
# Systolic/diastolic ratio by sex
get_ratios(svy, bpxsy1, bpxdi1, group = riagendr)
```

---

## 7. Weighted correlations — `get_corr()`

`get_corr()` estimates survey-weighted Pearson correlations between two or
more continuous variables. Confidence intervals use the Fisher Z
transformation, guaranteeing bounds in (−1, 1).

```{r corr-basic}
# Correlation between systolic BP and diastolic BP
get_corr(svy, c(bpxsy1, bpxdi1))
```

Pass three or more variables for a pairwise correlation matrix in long format:

```{r corr-multi}
get_corr(svy, c(bpxsy1, bpxdi1, bpxpls))
```

Each row is a unique variable pair. `r` is the weighted correlation, `p_value`
is from a two-tailed t-test, and `df = n - 2` are the degrees of freedom.

Switch to wide format for a more familiar correlation-matrix layout:

```{r corr-wide}
get_corr(svy, c(bpxsy1, bpxdi1, bpxpls), format = "wide")
```

---

## 8. Controlling uncertainty output

All analysis functions share a common `variance` argument. You can request
any combination of:

| Code   | What it returns                                     |
|--------|-----------------------------------------------------|
| `"se"` | Standard error                                      |
| `"ci"` | Confidence interval: `ci_low`, `ci_high`            |
| `"var"`| Variance (square of the SE)                         |
| `"cv"` | Coefficient of variation (SE / estimate)            |
| `"moe"`| Margin of error at `conf_level`                     |
| `"deff"` | Design effect (complex design variance / SRS variance) |

The `conf_level` argument controls the confidence level for `"ci"` and
`"moe"`. The default is `0.95`; for a 90% CI:

```{r variance-options}
get_means(svy, bpxsy1, variance = c("se", "ci", "moe"), conf_level = 0.90)
```

Set `variance = NULL` to suppress all uncertainty columns and return point
estimates and sample counts only.

### Estimated population counts

Add `n_weighted = TRUE` to any function to include the estimated population
count — the sum of weights — alongside the unweighted sample count `n`.

```{r n-weighted}
get_means(svy, bpxsy1, n_weighted = TRUE)
```

### Tidyverse-compatible output names

By default surveycore uses its own column names (`mean`, `se`, `ci_low`,
`ci_high`, etc.). Set `name_style = "broom"` to use names compatible with
**broom** and **modelsummary** (`estimate`, `std.error`, `conf.low`,
`conf.high`):

```{r broom-style}
get_means(svy, bpxsy1, name_style = "broom")
```

---

## 9. Weighted quantiles — `get_quantiles()`

`get_quantiles()` estimates survey-weighted quantiles using the Woodruff (1952)
confidence interval method. Confidence intervals are derived by inverting the
weighted CDF rather than assuming normality, so they are generally **asymmetric**
around the estimate and always respect the range of the data.

```{r quantiles-basic}
# Quartiles and median of age (default probs = c(0.25, 0.5, 0.75))
get_quantiles(svy, ridageyr)
```

The `quantile` column uses probability labels: `p25`, `p50`, `p75`, etc.

### Choosing quantiles

Pass any numeric vector to `probs`. For the median alone:

```{r quantiles-median}
get_quantiles(svy, ridageyr, probs = 0.5)
```

For deciles of systolic blood pressure:

```{r quantiles-deciles}
get_quantiles(svy, bpxsy1, probs = seq(0.1, 0.9, 0.1))
```

### Grouped quantiles

```{r quantiles-grouped}
# Median and IQR of systolic BP by sex
get_quantiles(svy, bpxsy1, group = riagendr)
```

> **Note:** `"deff"` is always `NA` for quantiles — there is no closed-form SRS
> standard error for quantile estimates, so the design effect cannot be computed.

---

## 10. surveytidy integration

The **surveytidy** package provides dplyr verbs — `filter()`, `select()`,
`mutate()`, and `group_by()` — that work directly on survey design objects.
Install it with:

```r
pak::pak("jacobdennen/surveytidy")
```

```{r surveytidy-check, include=FALSE}
has_surveytidy <- requireNamespace("surveytidy", quietly = TRUE)
```

### Domain estimation with `filter()`

A common pattern in survey analysis is computing an estimate for a
*subpopulation* — adults aged 60 and over, a particular racial group, one
treatment arm of a study. **Do not physically subset the data frame before
building the design object.** Removing rows changes the cluster and stratum
structure and produces incorrect standard errors [@lumley2010, ch. 4].

Use `filter()` instead. It marks rows as outside the domain while keeping
them in the design object so that the cluster structure remains intact:

```{r surveytidy-filter, eval=has_surveytidy}
library(surveytidy)

# Mean systolic BP for adults aged 60 and older
# The full cluster/strata structure is preserved for correct SEs
svy |>
  filter(ridageyr >= 60) |>
  get_means(bpxsy1)
```

### Derived variables with `mutate()`

`mutate()` creates new columns that are available to subsequent analysis
functions while preserving the survey design:

```{r surveytidy-mutate, eval=has_surveytidy}
# Create a hypertension indicator and estimate its prevalence
svy |>
  mutate(hypertension = bpxsy1 >= 130 | bpxdi1 >= 80) |>
  get_freqs(hypertension)
```

### Grouped analysis with `group_by()`

`group_by()` sets persistent grouping that is applied by all downstream
analysis functions — useful when you want multiple estimates from the same
grouped design:

```{r surveytidy-groupby, eval=has_surveytidy}
# Compute both means and frequencies for each sex group
svy_by_sex <- svy |> group_by(riagendr)

get_means(svy_by_sex, bpxsy1)
get_freqs(svy_by_sex, ridreth3)
```

You can also use the `group` argument directly in each function call, which
is equivalent and does not require surveytidy:

```{r group-equivalent}
# These two calls produce identical output:
get_means(svy, bpxsy1, group = riagendr)
```

---

## Summary

| Function | Use for |
|---|---|
| `get_freqs()` | Categorical variables — weighted distributions, percentages |
| `get_means()` | Continuous variables — weighted means |
| `get_totals()` | Population counts or aggregates — weighted sums |
| `get_ratios()` | Ratios of two weighted totals |
| `get_corr()` | Pairwise Pearson correlations |
| `get_quantiles()` | Weighted quantiles and median — Woodruff CIs |

All functions:
- Return a tibble subclass ready for further analysis or display
- Accept a `group` argument for subgroup estimates
- Accept a `variance` argument to control which uncertainty columns appear
- Handle all five survey design classes: `survey_taylor`, `survey_replicate`,
  `survey_srs`, `survey_twophase`, and `survey_calibrated`

---

## References
